Title Goes Here

### 2021-03-30 analysis/paper/paper.Rmd
Re-agrega introducción y conclusión

>>>>

One typical way of describing the zonally asymmetric circulation is by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height at each latitude [e.g. @vanloon1972;@trenberth1980]. These methodologies rely on the assumption that the circulation can be meaningfully understood in terms of zonal waves of constant amplitude along a latitude circle. This is not valid for meridionally propagating waves or zonal waves with localised amplitudes. In the case of the wave 3, for example, @trenberth1985 observed that wave 3 played a role in blocking events, but that in many cases, this was due to increased amplitude of a longitudinally localised wave train and not a hemispheric-scale zonal wave 3.

Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis). Within the EOF framework, the Southern Annular Mode appears as the leading mode of variability of the Southern Hemisphere (SAM) [@fogt2020] followed by the two Pacific–South American Patterns (PSA) [@mo2001]. The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector. The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula. These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.

EOFs are more flexible in the sense that they can characterise a richer set of spatial patterns, but they are restricted to standing oscillation modes and have difficulty dealing with propagating or phase-varying modes such as zonal waves. A single EOF can also represent a mixture of two or more physical modes.

The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs. @irving2016 characterises the Pacific–South American Pattern in a Fourier framework. Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator. On this new projection, they could identify the PSA using Fourier decomposition.

Finally, another, hybrid, methodology consist on identifying particular features of interest and creating indices using simple methods such as averages and differences. The Marshal SAM Index [@marshall2003], for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following @gong1999. @raphael2004 used a similar methodology to describe the wave 3 in the Southern Hemisphere. Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean. This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field. Similarly, @hobbs2010 posit that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.

These "feature-based" methods are grounded on other methods to identify the interesting features and can be useful to characterise features that are not readily apparent with Fourier or EOF methods but that nonetheless interesting or important. The use of data of single points, which makes them suitable for extending into the data-poor past also makes them not very robust.

In this paper we propose using complex Empirical Orthogonal Functions [@horel1984] as a robust extension of EOF analysis that is capable of describing rich structures with meridional and zonal propagation, and amplitude modulation such as the PSA. We apply this method to characterise the Springtime zonally asymmetric circulation of the extratopical Southern Hemisphere.
